---
title: "marine-genomics - Workshop 4"
output: html_document
date: "2026-07-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup

Create a new github repository called marine-genomics.
Create a project that is linked to github repository
Create RMarkdown document

## Assembly

Step 1. Create a working directory called flye and then make a symbolic link to the bbd read data

```bash
mkdir flye
ln -s /pvol/data/metagenome_assembly/bbd_0.1.fastq.gz .
```

Step 2. Make a shell script called run_flye_bbd.sh and include the following

```
#!/bin/bash
#SBATCH --time=720
#SBATCH --ntasks=12 --mem=10gb

echo "Starting flye bbd in $(pwd) at $(date)"

flye --nano-hq ../bbd_0.1.fastq.gz --threads 12 --out-dir bbd --meta

echo "Finished flye bbd in $(pwd) at $(date)"
```

Step 3. Submit run_flye_bbd.sh using sbatch

```bash
sbatch run_flye_bbd.sh
```

## Binning with metabat

Step 1. Create directory called metabat

```bash
mkdir metabat
cd metabat
```

Step 2. Extract depth information
```bash
cat ../flye/bbd/assembly_info.txt  | grep -v '#' | awk '{OFS="\t";print $1,$3}' > depths_bbd.txt
```

Step 3. Reorder contigs based on depth
```bash
cat depths_bbd.txt | awk '{print $1}' | xargs samtools faidx ../flye/bbd/assembly.fasta > assembly_bbd_ordered.fasta
```

Step 4. Run metabat
```bash
metabat2 -i assembly_bbd_ordered.fasta -a depths_bbd.txt --cvExt -o bins_bbd/metabat --seed 5
```

## Check bin quality with checkm

Step 1. Create directory called checkm

```bash
mkdir checkm
cd checkm
```

Step 2. Find the number of the largest contiguous in the bbd assembly (will need to manually add this contig because it usually isn't included in the bins from metabat)

```bash
cat ../flye/bbd/assembly.fasta | bioawk -c fastx '{print length($seq),$name}' | sort -n
```
contig 128 is the largest

Step 3. Create directory called bins_bbd and copy bins_bbd from metabat

```bash
mkdir bins_bbd
cp ../metabat/bins_bbd/*.fa bins_bbd/
samtools faidx ../flye/bbd/assembly.fasta contig_128 > bins_bbd/contig_128.fa
```

Step 4. Create a slurm script for checkm called run_checkm_bbd.sh in the checkm directory

```
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=6 --mem=40gb

echo "Starting checkm in $(pwd) at $(date)"

shopt -s expand_aliases
alias checkm='apptainer run -B /pvol/:/pvol /pvol/data/sif/checkm.sif checkm'

checkm lineage_wf bins_bbd out -t 4 -x fa -f checkm_results_bbd.txt

echo "Finished checkm in $(pwd) at $(date)"
```
Troubleshooting insight: Had to change memory from 10gb to 40gb due to the server being overloaded

Step 5. Submit run_checkm_bbd.sh using sbatch

```bash
sbatch run_checkm_bbd.sh
```

## Select bins based on checkm results

Step 1. Look at results from checkm analysis using head
```bash
head checkm_results_bbd.txt 
```

Step 2. Show a few more lines using head -n 20 
```bash
head -n 20 checkm_results_bbd.txt
```

Step 3. Use checkm completeness greater than 15% using awk to extract the bin IDs
```bash
cat checkm_results_bbd.txt | awk '$13>15'
```

Step 4. Change it from the entire line to just the bin ID using the print command
```bash
cat checkm_results_bbd.txt | awk '$13>15{print $1}'
```

Step 5. Send output to a file so it can be used in other steps
```bash
cat checkm_results_bbd.txt | awk '$13>15{print $1}' > bbd_topbins.txt
```

## Perform taxonomic classification with gtdbtk

Step 1. Create working directory called gtdbtk
```bash
mkdir gtdbtk
cd gtdbtk
```

Step 2. Create a subfolder within gtdbtk called genomes (this is where a copy of each of the top bins will go)
```bash
mkdir genomes
```

Step 3. Build the command and get the loop working
```bash
while read bin;do echo $bin;done < ../checkm/bbd_topbins.txt
```

Step 4. Use echo $bin to print the name of the bins. Turn the bin name into a path specifying the fasta file for the bin. Every bin will need to prepend the path to its folder ../metabat/bins_bbd/ and then append the file extension .fa
```bash
while read bin;do echo "../metabat/bins_bbd/${bin}.fa";done < ../checkm/bbd_topbins.txt
```

Step 5. Perform the copy from checkm to metabat
```bash
while read bin;do cp "../metabat/bins_bbd/${bin}.fa" genomes;done < ../checkm/bbd_topbins.txt
```
Troubleshooting insight: Had an error that said contig_128 not found. Had to copy contig_128 from checkm to metabat

Step 6. Check that it worked correctly by listing files within the genomes directory (should see one file for each of the bins)
```bash
ls genomes
```

Step 7. Create slurm script in the gtdbtk directory called run_gtdbtk.sh
```
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=40gb

echo "Starting gtdbtk in $(pwd) at $(date)"

shopt -s expand_aliases
alias gtdbtk='apptainer run -B /pvol:/pvol,/pvol/data/mg/gtdbtkdata/release207_v2/:/refdata /pvol/data/sif/gtdbtk.sif'

gtdbtk classify_wf -x fa --cpus 2 --genome_dir genomes --out_dir out --skip_ani_screen

echo "Finished gtdbtk in $(pwd) at $(date)"
```
Troubleshooting insight: Had to change memory from 24gb to 40gb due to the server being overloaded

Step 8. Submit run_gtdbtk.sh using sbatch
```bash
sbatch run_gtdbtk.sh
```
Step 9. Install tidyverse package
```bash
library(tidyverse)
```

Step 10. Check gtdbtk.bac120.summary.tsv to see the classifications
```{r}
gtdbtk_result <- read_tsv("gtdbtk.bac120.summary.tsv")
```

## Find and annotate genes with prokka

Step 1. 


